Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2015-12-10 paper.tex
Updated paper

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-40$~nT.  The second is an increase in $\rho_{eq}$ above 40~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using (2x ?) the standard deviation of the number of values used in computing the median divided by the square root of the number of values used.

@@

@@

A total of 669 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

\subsection{Mass Density Events}
Figure \ref{Storms}b shows this same algorithm, but looking for a rise in mass density over a value of $40$~amu/cm$^3$. This results in 130 events with a mean duration of 32 hours and a median duration of 17 hours, marked in red on the left figure.

This shows that when using all data for mass density derived events, almost no significant changes can be seen around event onset.

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold greater than $70$g/cm$^3$ results in the left and right sides of Figure \ref{Mspec} respectively.

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80$~nT, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80$~nT that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80$~nT, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80$~nT that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

@@

\section{F10.7 dependence}
In an effort to analyze the dependence of $\rho_{eq}$ on $F_{10.7}$, a few tests were performed. \cite{Takahashi2010} mention a strong correlation between the two. The long term correlation could be a bias for $D_{st}$'s effects, so a linear model was created, recreating mass density purely from $F_{10.7}$ in the form of $\rho_{eq}(t)=A\cdot F_{10.7}(t)$. This re-created $\rho_{eq}$ shows around a 45\% correlation with the actual $\rho_{eq}$, suggesting a strong influence, while doing the same procedure with $D_{st}$ shows only a 20-25\% correlation.

@@

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass-tak.eps}
\caption{Events in the interval 1989-1991. Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAverages}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-day.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass-day.eps}
\caption{Events in the interval 1983-1991.  Upper panel: $D_{st}$ events. Lower panel: $\rho_{eq}$ events.}
\label{DailyAveragesAllData}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-mass.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50$~nT onset. (b) Same as (a) except around time intervals where mass density crossed above $40$~amu/cm$^3$.}
\label{Storms}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dd12.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-md12.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours. (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ stayed above $50$~amu/cm$^3$ for at least 12 hours.}
\label{Mspec}
\end{figure}
\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-m70.eps}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT (b) Same as Figure \ref{Storms}(b) except for constrain that $\rho_{eq}$ crossed above $70$~amu/cm$^3$.}
\label{Dspec}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-diffden-10amu.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-diffden-30percent.eps}
\caption{(a) diff($\rho_{eq}$) $>$ 10$\frac{amu}{hour}$. (b) diff($\rho_{eq}$) $>$ 30$\frac{\%}{hour}$}
\label{rhochange}
\end{figure}
\clearpage

@@

\begin{figure}[tp]
\centering
\includegraphics[scale=0.40]{paperfigures/blockmedian.eps}
\includegraphics[scale=0.40]{paperfigures/blockmedian-prenoon.eps}
\caption{Comparing median values calculated as median of all 1-hour points in the 12 day block, and as the daily median of all hourly-mediated events. (b) Same, but only events starting between 0-12 UTC}
\label{blockmedian}
\end{figure}
\clearpage

\begin{figure}[tp]
\centering
\includegraphics[scale=0.9]{paperfigures/allstorms.eps}
\caption{Comparing all events to their median and $1\sigma$ range}
\label{allstorms}
\end{figure}
\clearpage

\begin{figure}[tp]
\centering
\includegraphics[scale=0.5]{paperfigures/nansbyhour.eps}
\includegraphics[scale=0.5]{paperfigures/nansbyhour_storm.eps}
\caption{(a) Number of NaN points per hour of observation in the total data set (b) $\rho_{eq}$ data availability relative to average event hour in events where $D_{st}<-80$~nT.}
\label{nanperhour}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1983-1991.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES7-1987-1991.eps}
\caption{GOES Satellite 6 and 7}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES2-1981-1983.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES5-1983-1987.eps}
\caption{GOES Satellite 2 and 5}
\end{figure}
\clearpage

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1989-1991.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst80.eps}
\caption{Top: From 89-91 (compare to Takahashi 2010). Bottom:  $D_{st}<-80$~nT.}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/AEbyhour.eps}
\caption{$AE$ events by their hour of onset}
\end{figure}
\clearpage

@@

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1983-1991.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES7-1987-1991.eps}
\caption{Comparing GOES 6 to GOES 7}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1983-1987.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1987-1991.eps}
\caption{GOES 6 from 83-87 and 87-91}
\end{figure}

>>>>

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-50$~nT.  The second is an increase in $\rho_{eq}$ above 40~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of the number of values used in computing the median divided by the square root of the number of values used.

@@

In the lower panel of \ref{DailyAverages} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire run of GOES 6 over 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

It was found that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancements to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

@@

A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-50$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold less than $-80$nT results in Figure \ref{Dspec}. A spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 211 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80$~nT that have existing mass density data around onset, so an analysis of this particular combination of factors can't be made.

@@

\section{F10.7 dependence}
In an effort to analyze the dependence of $\rho_{eq}$ on $F_{10.7}$, a few tests were performed. \cite{Takahashi2010} mention a strong correlation between the two. The long term correlation could be a bias for $D_{st}$'s effects, so a linear model was created, recreating mass density purely from $F_{10.7}$ in the form of $\rho_{eq}(t)=a\cdot F_{10.7}(t)$. This re-created $\rho_{eq}$ shows around a 45\% correlation with the actual $\rho_{eq}$, suggesting a strong influence, while doing the same procedure with $D_{st}$ shows only a 20-25\% correlation.

@@

Looking at GOES 2,5,6, and 7 (Figures \ref{GOES67,GOES25}) shows that this post-onset $D_{st}$ spike happens in almost every case for periods of high $F_{10.7}$.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-day.eps}
\caption{Events in the interval 1989-1991. Upper panel: $D_{st}$ events. Lower panel: Events in the interval 1983-1991.}
\label{DailyAverages}
\end{figure}

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dd12.eps}
\caption{(a) Average of solar wind and near-Earth measurements around the time $D_{st}$ crossed below $-50$~nT onset. (b) Same as (a) except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours.}
\label{Storms}
\end{figure}

\clearpage

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-d80.eps}
\rule[1ex]{5cm}{1pt}
\caption{(a) Same as Figure \ref{Storms}(a) except with constraint that $D_{st}$ crossed below $-80$~nT }
\label{Dspec}
\end{figure}

@@

\clearpage

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1983-1991.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES7-1987-1991.eps}
\caption{GOES Satellite 6 and 7}
\label{GOES67}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES2-1981-1983.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES5-1983-1987.eps}
\caption{GOES Satellite 2 and 5}
\label{GOES25}
\end{figure}
\clearpage

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1989-1991.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst80.eps}
\caption{Top: GOES 6 from 89-91 (compare to Takahashi 2010). Bottom:  $D_{st}<-80$~nT.}
\end{figure}

\begin{figure}[htp!]
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1983-1987.eps}
\includegraphics[scale=0.45]{paperfigures/HighLowF107rhoeq-Dst50-GOES6-1987-1991.eps}
\caption{GOES 6 from 83-87 and 87-91}
\end{figure}
\clearpage

@@

@@
